Semantics for the $ \lambda $-calculus

### 2024-04-08 report/chapters/previous-work.tex
Show that R does not have coproducts

@@

Note that this category has many (isomorphic, but not equal for nontrivial $ L $) terminal objects, given by $ I_c := \lambda x_1, \iota_{0, 1}(c) : \R $ for any $ c: L_0 $. These are terminal because for $ f: A \to I_c $, we have $ f = I_c \circ f = I_c $. Note that $ \varphi(I_c)_n = \{ \iota_{0, n}(c) \} $. We will choose $ I = \lambda x_1 x_2, x_2 : \R $ as our main example of the terminal object.

We might wonder whether the category also has an initial object $ O $. However, for any $ c: L_0 $, we would have a morphism (with the same term as the terminal object)
\[ \lambda x_1, \iota_{0, 1}(c) : A \to U, \]
so if $ L $ is nontrivial, $ \R $ has no initial object.

The category has binary products with projections and product morphisms
\[ A_1 \times A_2 = \langle p_1, p_2 \rangle, \quad p_i = A_i \circ \pi_i \quad \text{and} \quad \langle f, g \rangle. \]
Recall that for any object $ A $, we have $ A = \id A $, which for $ A_1 \times A_2 $ is $ \langle p_1, p_2 \rangle $ by the universal property of the product, which explains why the product is of this form.

The category also has exponential objects
\[ C^B = \lambda x_1, C \circ x_1 \circ B \]
with evaluation morphism $ \epsilon_{BC}: C^B \times B \to C $ given by
\[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
which is universal because we can lift a morphism $ f: \R(A \times B, C) $ to a morphism $ \psi(f): \R(A, C^B) $ given by
\[ \psi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
Note that for $ g: \R(A, C^B) $, the inverse $ \psi^{-1}(g) $ is given by
\[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle = \lambda x_1, g (\pi_1 x_1) (\pi_2 x_1): \R(A \times B, C). \]
Also note that
\[ \lambda x_1, C \circ x_1 \circ B = C^B = \id{C^B} = \psi(\epsilon_{BC}). \]
The first representation shows that $ \varphi(C^B)_0 = \R(B, C) $, as we would expect.

@@

>>>>

\begin{tabular}{|c|c|c|}\hline
  $ K $ & $ \lambda xy.x $ & constant \\\hline
  $ S $ & $ \lambda xyz.xz(yz) $ & \\\hline
  $ I $ & $ \lambda x.x = SKK $ & identity \\\hline
  $ Y $ & $ \lambda f.(\lambda x.f (xx))(\lambda x.f (xx)) $ & ﬁxpoint \\\hline
  $ \bot $ & $ (\lambda x.xx)(\lambda x.xx) = (SII)(SII) = YI $ & bottom \\\hline
  $ P $ & $ \lambda f gx.g(f x) = S(K(S(S(KS)K)))K $ & composition \\\hline
  $ Q $ & $ \lambda f gax.ga(f ax) $ & ﬁbred composition \\\hline
  $ \langle \rangle $ & $ \lambda xyz.zxy $ & pairing \\\hline
  $ 0 $ & $ \lambda xy.x = K $ & left projection \\\hline
  $ 1 $ & $ \lambda xy.y = KI $ & right projection \\\hline
\end{tabular}

@@

Note that $ \R $ has many (isomorphic, but not equal for nontrivial $ L $) terminal objects, given by $ I_c := \lambda x_1, \iota_{0, 1}(c) : \R $ for any $ c: L_0 $. These are terminal because for $ f: A \to I_c $, we have $ f = I_c \circ f = I_c $. Note that $ \varphi(I_c)_n = \{ \iota_{0, n}(c) \} $. We will choose $ I = \lambda x_1 x_2, x_2 : \R $ as our main example of the terminal object.

We might wonder whether $ \R $ also has an initial object $ O $. However, for any $ c: L_0 $, we would have a morphism (with the same term as the terminal object)
\[ \lambda x_1, \iota_{0, 1}(c) : A \to U, \]
so if $ L $ is nontrivial, $ \R $ has no initial object.

$ \R $ has binary products with projections and product morphisms
\[ A_1 \times A_2 = \langle p_1, p_2 \rangle, \quad p_i = A_i \circ \pi_i \quad \text{and} \quad \langle f, g \rangle. \]
Recall that for any object $ A $, we have $ A = \id A $, which for $ A_1 \times A_2 $ is $ \langle p_1, p_2 \rangle $ by the universal property of the product, which explains why the product is of this form.

$ \R $ also has exponential objects
\[ C^B = \lambda x_1, C \circ x_1 \circ B \]
with evaluation morphism $ \epsilon_{BC}: C^B \times B \to C $ given by
\[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
which is universal because we can lift a morphism $ f: \R(A \times B, C) $ to a morphism $ \psi(f): \R(A, C^B) $ given by
\[ \psi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
Note that for $ g: \R(A, C^B) $, the inverse $ \psi^{-1}(g) $ is given by
\[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle = \lambda x_1, g (\pi_1 x_1) (\pi_2 x_1): \R(A \times B, C). \]
Also note that
\[ \lambda x_1, C \circ x_1 \circ B = C^B = \id{C^B} = \psi(\epsilon_{BC}). \]
The first representation shows that $ \varphi(C^B)_0 = \R(B, C) $, as we would expect.

@@

Taylor notes (\cite{taylor}, Section 1.5) that the objects in $ \R $ have very strong properties with respect to fixpoints. One of the properties also arises via Lawvere's fixed point theorem (\cite{lawvere-fixpoints}, page 136): For all $ B: \R $, since $ B^U $ is a retract of $ U $, every endomorphism $ f: B \to B $ has a fixpoint. That is, there exists $ s: I \to B $ such that $ f \circ s = s $. Working out the proof even yields an explicit term:
\[ s = \lambda i, (\lambda x, f (x x)) (\lambda x, f (x x)). \]
Indeed, $ s = \lambda i, f ((\lambda x, f (x x)) (\lambda x, f (x x))) = f \circ s $, and $ B \circ s = B \circ f \circ s = s = s \circ I $, so $ s : \R(I, B) $.

From this, Taylor deduces (\cite{taylor}, \S 1.5.12) that $ \R $ does not have all coproducts if $ L $ is nontrivial, because suppose that $ \R $ has all coproducts. Then $ B = I + I $ is a `boolean algebra object': We define $ \bot, \top: I \to B $ to be the injections on the left and right components. Since $ \R $ is cartesian closed, binary products distribute over binary coproducts, so we have $ B \times B \cong ((I \times I) + (I \times I)) + ((I \times I) + (I \times I)) $, and note that $ I \times I \cong I $. Using the universal property of the coproduct, we can define $ \lnot: B \to B $ componentwise as $ \lnot = [\top, \bot] $, the coproduct arrow
\begin{center}
  \begin{tikzcd}[row sep = .3in]
    I \arrow[r, "\bot"] \arrow[rd, "\top"'] & I + I \arrow[d, dashed, "{[\top, \bot]}"description] & I \arrow[l, "\top"'] \arrow[ld, "\bot"]\\
    & I + I &
  \end{tikzcd}
  % \begin{tikzcd}[row sep = .3in]
  %   I \arrow[rrrd, "\bot"'] \arrow[r, "\bot"] & I + I \arrow[rrd, "{[\bot, \bot]}" description, dashed] \arrow[rr, bend left] & I \arrow[rd, "\bot"] \arrow[l, "\top"'] & I + I + I + I \arrow[d, dashed, "{[[\bot, \bot], [\bot, \top]]}" description] & I \arrow[ld, "\bot"'] \arrow[r, "\bot"] & I + I \arrow[lld, "{[\bot, \top]}" description, dashed] \arrow[ll, bend right] & I \arrow[llld, "\top"] \arrow[l, "\top"'] \\
  %   & & & I + I & & &
  % \end{tikzcd}
\end{center}
Note that by the definition of $ \lnot $, $ \lnot \circ \bot = \top $. Similarly, we define $ \land, \lor : B \times B \to B $ as $ \land = [[\bot, \bot], [\bot, \top]] $ and $ \lor = [[\bot, \top], [\top, \top]] $. Using the same universal property, we can also verify some properties of $ \bot, \top, \land $ and $ \lnot $, like the fact that the following diagrams commute:
\begin{center}
  \begin{tikzcd}[column sep = .5in]
    B \arrow[rd, "{\id B}"] \arrow[r, "{\langle \id B, \id B \rangle}"] & B \times B \arrow[d, "\land"]\\
    & B
  \end{tikzcd}
  \begin{tikzcd}[column sep = .5in]
    B \arrow[d, "!"] \arrow[r, "{\langle \lnot, \id B \rangle}"] & B \times B \arrow[d, "\land"]\\
    I \arrow[r, "\bot"] & B
  \end{tikzcd}
\end{center}
for $ !: B \to I $ the terminal projection. We do this by checking, for example, that $ \id B \circ \top = \land \circ \langle \id B, \id B \rangle \circ \top $ and $ \id B \circ \bot = \land \circ \langle \id B, \id B \rangle \circ \bot $. Now, as mentioned above, every endomorphism has a fixed point. In particular, we have some $ \star: I \to B $ such that $ \lnot \circ \star = \star $. Now, note that
\[ \star = \land \circ \langle \id B, \id B \rangle \circ \star = \land \circ \langle \star, \star \rangle = \land \circ \langle \star, \lnot \circ \star \rangle = \land \circ \langle \id B, \lnot \rangle \circ \star = \bot \circ ! \circ \star = \bot \]
and then
\[ \bot = \star = \lnot \circ \star = \lnot \circ \bot = \top. \]
Now, for any object $ A: \R $ and any two global elements $ f, g: I \to A $, we have the following diagram:
\begin{center}
  \begin{tikzcd}
    I \arrow[r, shift left, "\bot"] \arrow[r, shift right, "\top"'] \arrow[rr, bend left, "f"] \arrow[rr, bend right, "g"'] & I + I \arrow[r, "{[f, g]}"] & A
  \end{tikzcd}
\end{center}
and we have
\[ f = [f, g] \circ \bot = [f, g] \circ \top = g. \]
In particular, for any two objects $ A, A^\prime: \R $,
\[ \R(A^\prime, A) \simeq \R(I \times A^\prime, A) \simeq \R(I, A^{A^\prime}) \simeq \{ \star \}, \]
so $ \R $ the trivial category and $ L $ is trivial.
